---
title: "Regional association plotting of 11 loci associated with CAC."
author: "[Sander W. van der Laan, PhD](https://swvanderlaan.github.io) | @swvanderlaan | s.w.vanderlaan@gmail.com"
date: "`r Sys.Date()`"
output:
  html_notebook:
    cache: yes
    code_folding: hide
    collapse: yes
    df_print: paged
    fig.align: center
    fig_caption: yes
    fig_height: 6
    fig_retina: 2
    fig_width: 7
    highlight: tango
    theme: lumen
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
mainfont: Arial
subtitle: "A 'druggable-MI-targets' project"
editor_options:
  chunk_output_type: inline
---

```{r global_options, include = FALSE}
# further define some knitr-options.
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'Figures/', 
                      wwarning = TRUE, # show warnings during codebook generation
  message = TRUE, # show messages during codebook generation
  error = TRUE, # do not interrupt codebook generation in case of errors,
                # usually better for debugging
  echo = TRUE,  # show R code
                      eval = TRUE)
ggplot2::theme_set(ggplot2::theme_minimal())
pander::panderOptions("table.split.table", Inf)
```

# Setup
We will clean the environment, setup the locations, define colors, and create a datestamp.

_Clean the environment._
```{r echo = FALSE}
rm(list = ls())
```

_Set locations and working directories..._
```{r LocalSystem, echo = FALSE}
### Operating System Version
### MacBook Pro
ROOT_loc = "/Users/swvanderlaan/OneDrive - UMC Utrecht"
# STORAGE_loc = "/Volumes/LaCie/"
STORAGE_loc = "/Users/swvanderlaan/"

### MacBook Air
# ROOT_loc = "/Users/slaan3/OneDrive - UMC Utrecht"
# STORAGE_loc = "/Volumes/LaCie/"
# STORAGE_loc = "/Users/slaan3/"

GENOMIC_loc = paste0(ROOT_loc, "/Genomics")
AEDB_loc = paste0(GENOMIC_loc, "/Athero-Express/AE-AAA_GS_DBs")
LAB_loc = paste0(GENOMIC_loc, "/LabBusiness")

PLINK_loc=paste0(STORAGE_loc,"/PLINK")
AEGSQC_loc =  paste0(PLINK_loc, "/_AE_ORIGINALS/AEGS_COMBINED_QC2018")
MICHIMP_loc=paste0(PLINK_loc,"/_AE_ORIGINALS/AEGS_COMBINED_EAGLE2_1000Gp3v5HRCr11")

GWAS_loc=paste0(PLINK_loc,"/_GWAS_Datasets/_CHARGE_CAC")

PROJECT_loc = paste0(PLINK_loc, "/analyses/consortia/CHARGE_1000G_CAC")

# use this if there is relevant information here.
TARGET_loc = paste0(PROJECT_loc, "/targets")

### SOME VARIABLES WE NEED DOWN THE LINE
TRAIT_OF_INTEREST = "CAC" # Phenotype
PROJECTNAME = "CAC"

cat("\nCreate a new analysis directory...\n")
ifelse(!dir.exists(file.path(PROJECT_loc, "/",PROJECTNAME)), 
       dir.create(file.path(PROJECT_loc, "/",PROJECTNAME)), 
       FALSE)
ANALYSIS_loc = paste0(PROJECT_loc,"/",PROJECTNAME)

ifelse(!dir.exists(file.path(ANALYSIS_loc, "/PLOTS")), 
       dir.create(file.path(ANALYSIS_loc, "/PLOTS")), 
       FALSE)
PLOT_loc = paste0(ANALYSIS_loc,"/PLOTS")

ifelse(!dir.exists(file.path(PLOT_loc, "/QC")), 
       dir.create(file.path(PLOT_loc, "/QC")), 
       FALSE)
QC_loc = paste0(PLOT_loc,"/QC")

ifelse(!dir.exists(file.path(ANALYSIS_loc, "/OUTPUT")), 
       dir.create(file.path(ANALYSIS_loc, "/OUTPUT")), 
       FALSE)
OUT_loc = paste0(ANALYSIS_loc, "/OUTPUT")

ifelse(!dir.exists(file.path(ANALYSIS_loc, "/BASELINE")), 
       dir.create(file.path(ANALYSIS_loc, "/BASELINE")), 
       FALSE)
BASELINE_loc = paste0(ANALYSIS_loc, "/BASELINE")

ifelse(!dir.exists(file.path(PROJECT_loc, "/SNP")), 
       dir.create(file.path(PROJECT_loc, "/SNP")), 
       FALSE)
SNP_loc = paste0(PROJECT_loc, "/SNP")

setwd(paste0(PROJECT_loc))
getwd()
list.files()

```

_... a package-installation function ..._
```{r}
source(paste0(PROJECT_loc, "/scripts/functions.R"))
```


_... and load those packages._
```{r loading_packages, message=FALSE, warning=FALSE}
install.packages.auto("readr")
install.packages.auto("optparse")
install.packages.auto("tools")
install.packages.auto("dplyr")
install.packages.auto("tidyr")
install.packages.auto("naniar")

# To get 'data.table' with 'fwrite' to be able to directly write gzipped-files
# Ref: https://stackoverflow.com/questions/42788401/is-possible-to-use-fwrite-from-data-table-with-gzfile
# install.packages("data.table", repos = "https://Rdatatable.gitlab.io/data.table")
library(data.table)

install.packages.auto("tidyverse")
install.packages.auto("knitr")
install.packages.auto("DT")
install.packages.auto("eeptools")

install.packages.auto("haven")
install.packages.auto("tableone")

install.packages.auto("BlandAltmanLeh")

# Install the devtools package from Hadley Wickham
install.packages.auto('devtools')

# for plotting
install.packages.auto("pheatmap")
install.packages.auto("forestplot")
install.packages.auto("ggplot2")

install.packages.auto("ggpubr")

install.packages.auto("UpSetR")

devtools::install_github("thomasp85/patchwork")

# For regional association plots
install.packages.auto("devtools") 
library(devtools) 
install_github("oliviasabik/RACER") 

```

_We will create a datestamp and define the Utrecht Science Park Colour Scheme_.
```{r Setting: Colors}

Today = format(as.Date(as.POSIXlt(Sys.time())), "%Y%m%d")
Today.Report = format(as.Date(as.POSIXlt(Sys.time())), "%A, %B %d, %Y")

### UtrechtScienceParkColoursScheme
###
### WebsitetoconvertHEXtoRGB:http://hex.colorrrs.com.
### Forsomefunctionsyoushoulddividethesenumbersby255.
###
###	No.	Color			      HEX	(RGB)						              CHR		  MAF/INFO
###---------------------------------------------------------------------------------------
###	1	  yellow			    #FBB820 (251,184,32)				      =>	1		or 1.0>INFO
###	2	  gold			      #F59D10 (245,157,16)				      =>	2		
###	3	  salmon			    #E55738 (229,87,56)				      =>	3		or 0.05<MAF<0.2 or 0.4<INFO<0.6
###	4	  darkpink		    #DB003F ((219,0,63)				      =>	4		
###	5	  lightpink		    #E35493 (227,84,147)				      =>	5		or 0.8<INFO<1.0
###	6	  pink			      #D5267B (213,38,123)				      =>	6		
###	7	  hardpink		    #CC0071 (204,0,113)				      =>	7		
###	8	  lightpurple	    #A8448A (168,68,138)				      =>	8		
###	9	  purple			    #9A3480 (154,52,128)				      =>	9		
###	10	lavendel		    #8D5B9A (141,91,154)				      =>	10		
###	11	bluepurple		  #705296 (112,82,150)				      =>	11		
###	12	purpleblue		  #686AA9 (104,106,169)			      =>	12		
###	13	lightpurpleblue	#6173AD (97,115,173/101,120,180)	=>	13		
###	14	seablue			    #4C81BF (76,129,191)				      =>	14		
###	15	skyblue			    #2F8BC9 (47,139,201)				      =>	15		
###	16	azurblue		    #1290D9 (18,144,217)				      =>	16		or 0.01<MAF<0.05 or 0.2<INFO<0.4
###	17	lightazurblue	  #1396D8 (19,150,216)				      =>	17		
###	18	greenblue		    #15A6C1 (21,166,193)				      =>	18		
###	19	seaweedgreen	  #5EB17F (94,177,127)				      =>	19		
###	20	yellowgreen		  #86B833 (134,184,51)				      =>	20		
###	21	lightmossgreen	#C5D220 (197,210,32)				      =>	21		
###	22	mossgreen		    #9FC228 (159,194,40)				      =>	22		or MAF>0.20 or 0.6<INFO<0.8
###	23	lightgreen	  	#78B113 (120,177,19)				      =>	23/X
###	24	green			      #49A01D (73,160,29)				      =>	24/Y
###	25	grey			      #595A5C (89,90,92)				        =>	25/XY	or MAF<0.01 or 0.0<INFO<0.2
###	26	lightgrey		    #A2A3A4	(162,163,164)			      =>	26/MT
###
###	ADDITIONAL COLORS
###	27	midgrey			#D7D8D7
###	28	verylightgrey	#ECECEC"
###	29	white			#FFFFFF
###	30	black			#000000
###----------------------------------------------------------------------------------------------

uithof_color = c("#FBB820","#F59D10","#E55738","#DB003F","#E35493","#D5267B",
                 "#CC0071","#A8448A","#9A3480","#8D5B9A","#705296","#686AA9",
                 "#6173AD","#4C81BF","#2F8BC9","#1290D9","#1396D8","#15A6C1",
                 "#5EB17F","#86B833","#C5D220","#9FC228","#78B113","#49A01D",
                 "#595A5C","#A2A3A4", "#D7D8D7", "#ECECEC", "#FFFFFF", "#000000")

uithof_color_legend = c("#FBB820", "#F59D10", "#E55738", "#DB003F", "#E35493",
                        "#D5267B", "#CC0071", "#A8448A", "#9A3480", "#8D5B9A",
                        "#705296", "#686AA9", "#6173AD", "#4C81BF", "#2F8BC9",
                        "#1290D9", "#1396D8", "#15A6C1", "#5EB17F", "#86B833",
                        "#C5D220", "#9FC228", "#78B113", "#49A01D", "#595A5C",
                        "#A2A3A4", "#D7D8D7", "#ECECEC", "#FFFFFF", "#000000")
### ----------------------------------------------------------------------------
```

# Introduction

We will parse the data to create regional association plots for each of the 11 loci. 

# Load data

We need to load the data first.
```{r}

gwas_sumstats_racer <- readRDS(file = paste0(OUT_loc, "/gwas_sumstats_racer.rds"))

```

# Regional association plotting

## Top 11 loci

We are interested in 11 top loci. 

```{r}
library(openxlsx)
variant_list <- read.xlsx(paste0(TARGET_loc, "/Variants.xlsx"), sheet = "TopLoci")

DT::datatable(variant_list)

```


Let's do some plotting.


```{r}
library(RACER)
# Make directory for plots
ifelse(!dir.exists(file.path(PROJECT_loc, "/RACER")), 
       dir.create(file.path(PROJECT_loc, "/RACER")), 
       FALSE)
RACER_loc = paste0(PROJECT_loc,"/RACER")

variants_of_interest <- c(variant_list$rsID)

for(VARIANT in variants_of_interest){
  cat(paste0("Getting data for ", VARIANT,".\n"))

  tempCHR <- subset(variant_list, rsID == VARIANT)[,5]
  tempSTART <- subset(variant_list, rsID == VARIANT)[,17]
  tempEND <- subset(variant_list, rsID == VARIANT)[,18]
  tempVARIANTnr <- subset(variant_list, rsID == VARIANT)[,1]

  cat("\nSubset required data.\n")
  temp <- subset(gwas_sumstats_racer, Chr == tempCHR & (Position >= tempSTART & Position <= tempEND))
  
  cat("\nFormatting association data.\n")
  temp_f = RACER::formatRACER(assoc_data = temp, chr_col = 3, pos_col = 4, p_col = 5)

  cat("\nGetting LD data.\n")
  temp_f_ld = RACER::ldRACER(assoc_data = temp_f, rs_col = 2, pops = "EUR", lead_snp = VARIANT)
  
  cat(paste0("\nPlotting region surrounding ", VARIANT," on ",tempCHR,":",tempSTART,"-",tempEND,".\n"))
  p1 <- singlePlotRACER2(assoc_data = temp_f_ld, 
                               chr = tempCHR, build = "hg19", 
                               plotby = "snp", snp_plot = VARIANT,
                               label_lead = TRUE)
  
  print(p1)
  cat(paste0("Saving image for ", VARIANT,".\n"))
  ggsave(filename = paste0(RACER_loc, "/", tempVARIANTnr, ".", Today, ".",VARIANT,".regional_assoc.png"), plot = last_plot())
  ggsave(filename = paste0(RACER_loc, "/", tempVARIANTnr, ".", Today, ".",VARIANT,".regional_assoc.pdf"), plot = last_plot())
  ggsave(filename = paste0(RACER_loc, "/", tempVARIANTnr, ".", Today, ".",VARIANT,".regional_assoc.ps"), plot = last_plot())
  
  rm(temp, p1,
     temp_f, temp_f_ld,
     tempCHR, tempSTART, tempEND,
     VARIANT, tempVARIANTnr)
  
}


```


## Additional regional plots

We want to create some regional association plots to combine with teh UCSC browser tracks, thus we need the exact same regions. 

```{r}
library(openxlsx)
add_list <- read.xlsx(paste0(TARGET_loc, "/Variants.xlsx"), sheet = "AdditionalPlots")

DT::datatable(add_list)

```

```{r}
# Make directory for plots
ifelse(!dir.exists(file.path(PROJECT_loc, "/RACER")), 
       dir.create(file.path(PROJECT_loc, "/RACER")), 
       FALSE)
RACER_loc = paste0(PROJECT_loc,"/RACER")

variants_of_interest <- c(add_list$rsID)


for(VARIANT in variants_of_interest){
  cat(paste0("Getting data for ", VARIANT,".\n"))

  tempCHR <- subset(add_list, rsID == VARIANT)[,4]
  tempSTART <- subset(add_list, rsID == VARIANT)[,5]
  tempEND <- subset(add_list, rsID == VARIANT)[,6]
  tempNAME <- subset(add_list, rsID == VARIANT)[,3]

  cat("\nSubset required data.\n")
  temp <- subset(gwas_sumstats_racer, Chr == tempCHR & (Position >= tempSTART & Position <= tempEND))
  
  cat("\nFormatting association data.\n")
  temp_f = RACER::formatRACER(assoc_data = temp, chr_col = 3, pos_col = 4, p_col = 5)

  cat("\nGetting LD data.\n")
  temp_f_ld = RACER::ldRACER(assoc_data = temp_f, rs_col = 2, pops = "EUR", lead_snp = VARIANT)
  
  cat(paste0("\nPlotting region surrounding ", VARIANT," on ",tempCHR,":",tempSTART,"-",tempEND,".\n"))
  
  p1 <- singlePlotRACER2(assoc_data = temp_f_ld, 
                               chr = tempCHR, build = "hg19", 
                               plotby = "coord", snp_plot = VARIANT,
                               start_plot = tempSTART, end_plot = tempEND,
                               label_lead = TRUE)
  
  print(p1)
  
  cat(paste0("Saving image for ", VARIANT,".\n"))
  ggsave(filename = paste0(RACER_loc, "/", tempNAME, ".", Today, ".",VARIANT,".",tempSTART,".",tempEND,".regional_assoc.png"), plot = last_plot())
  ggsave(filename = paste0(RACER_loc, "/", tempNAME, ".", Today, ".",VARIANT,".",tempSTART,".",tempEND,".regional_assoc.pdf"), plot = last_plot())
  ggsave(filename = paste0(RACER_loc, "/", tempNAME, ".", Today, ".",VARIANT,".",tempSTART,".",tempEND,".regional_assoc.ps"), plot = last_plot())
  
  rm(temp, p1,
     temp_f, temp_f_ld,
     tempCHR, tempSTART, tempEND,
     VARIANT, tempNAME)
  
}
```


# Session information

------

    Version:      v1.2.0
    Last update:  2022-01-28
    Written by:   Sander W. van der Laan (s.w.vanderlaan-2[at]umcutrecht.nl).
    Description:  Script to create plot regional association plots.
    Minimum requirements: R version 3.4.3 (2017-06-30) -- 'Single Candle', Mac OS X El Capitan
    
    Changes log
    * v1.2.0 Added in aditional regions.
    * v1.1.0 Created PNG and PDF of top loci regions.
    * v1.0.0 Initial version. 

------

```{r eval = TRUE}
sessionInfo()
```


# Saving environment
```{r Saving}

save.image(paste0(PROJECT_loc, "/",Today,".",PROJECTNAME,".RegionalAssociationPlots.RData"))
```


------
<sup>&copy; 1979-2022 Sander W. van der Laan | s.w.vanderlaan[at]gmail.com | [swvanderlaan.github.io](https://swvanderlaan.github.io).</sup>
------

  
